---
title: "Munging Enhancer Data"
author: "Coleman Breen"
date: "2022-12-13"
output: pdf_document
---

```{r}
library(data.table)
library(tidyverse)
```

```{r}
enhancer.df <- fread("../DataReference/enhancersFromFANTOM/human_permissive_enhancers_phase_1_and_2_expression_tpm_matrix.txt")

samples.df <- read_delim(
  "../DataReference/enhancersFromFANTOM/Human.sample_name2library_id.txt", 
  delim = "\t",
  col_names = c("description", "id"))

head(samples.df)

```



# Munging Enhancer Data

FANTOM provides enhancer calls for a ton of tissue types. If you want to look at specific tissue enhancers, you can use the FANTOM-UCSC integration tool. If you want to download all of the enhancers, you can do so. But then you lose tissue annotation.

It appears we have three individual samples for blood and one donor / one pool for brain. What does this mean?

```{r}
samples_of_interest <- c("CNhs11958",
                         "CNhs10617", "CNhs11796",
                         "CNhs12002", "CNhs10860")

samples.df %>%
  dplyr::filter(id %in% samples_of_interest)
```

How well do these assays agree with each other?

```{r}
round(cor(enhancer.df[ , ..samples_of_interest]), 2)
```

The first, fourth, and fifth samples are the bloods--all have correlations above 0.74. 
