---
title: "CTCF Analysis"
author: "Coleman Breen"
date: "2022-11-28"
output: pdf_document
---

```{r}
library(tidyverse)
library(data.table)
library(poolr)
library(GenomicRanges)
```

```{r}
pvals <- fread("../DataRaw/pvals.bed")
ctcfs <- fread("../DataReference/GRCh38-CTCF.bed", 
               col.names = c("chr", "start", "end", "V1", "V2", "description"),              ) %>% dplyr::select(c(chr, start, end, description))
```


# CTCF 

```{r}
ctcfs %>% 
  dplyr::mutate(width = end - start) %>% 
  ggplot(aes(x = width)) +
  geom_histogram(bins = 50) +
  theme_minimal() +
  ggtitle("Width of CTCF-binding motifs in ENCODE database")
```

# Overlap 

```{r}
pvals.gr <- makeGRangesFromDataFrame(pvals, 
                                     keep.extra.columns = T, 
                                     ignore.strand = T, 
                                     starts.in.df.are.0based = T
                                     )


ctcf.gr <- makeGRangesFromDataFrame(ctcfs, 
                                    keep.extra.columns = T, 
                                    ignore.strand = T, 
                                    starts.in.df.are.0based = T
                                    )

# Overlap
overlaps <- findOverlaps(pvals.gr, ctcf.gr)

```

## How many CpGs per CTCF?

```{r}

n.cpgs <- length(queryHits(overlaps))
n.unique.cpgs <- length(unique(queryHits(overlaps)))

print(paste("Number of times a CpG overlaps a CTCF domain:", n.cpgs))
print(paste("Number of unique CpGs in at least one CTCF domain:", n.unique.cpgs))

```

There are ~2.8 million instances where a CpG overlaps a CTCF. Are CTCF domains overlapping? How many CpGs per CTCF? How mnay CTCFs without any CpGs?

```{r}
print(paste("Number of CTCF domains to start:", length(ctcf.gr)))

print(paste("Number of CTCF domains with at least one CpG:",
            length(unique(subjectHits(overlaps)))))
```
Of the 450k CTCF binding domains, we lose about 50k because they do not have any CpGs.

```{r}
data.frame(overlaps) %>% 
  group_by(subjectHits) %>%
  summarize(N.CpGs = n()) %>% 
  ggplot(aes(x = N.CpGs)) +
  geom_histogram(bins = 50) +
  theme_minimal() +
  ggtitle("Majority of CTCF-binding domains have fewer than 10 CpGs")
```

# Do the p-values 

```{r}
ix <- queryHits(overlaps)

pvals[ix, ] %>% 
  ggplot(aes(x = p.corrected)) +
  geom_histogram() +
  theme_minimal()
```

Low p-value sites are somewhat absent from CTCF-binding domains. However, if there are, say, 10 out of 10 CpGs which are DMPs in one CTCF, that might mean there is something interesting nearby.

Show the p-values associated with one CTCF region (arbitrarily picked 71).

```{r}

pvals.ix <- queryHits(overlaps)
ctcf.ix <- subjectHits(overlaps)

fisher_wrapper <- function(z){
   # Which CpGs are in the zth ctcf
  keepix <- (ctcf.ix == z)
  pp <- pvals.gr[pvals.ix[keepix]]$p.corrected
  
  # Run Fisher's method to combine p-values
  fisher.out <- fisher(pp)
  fisher.out$p
}

z_range <- unique(ctcf.ix)

```

Now do the same thing for all 400k regions.

```{r}

ofile <- "../DataDerived/ctcf.fisher.RData"

if (file.exists(ofile)){
  load(ofile)
  
} else {
  print("Computing Fisher p-values, this may take 10-15 mins")
  
  # Run the Fisher routine on all the CTCF binding domains
  ctcf.p <- 
    do.call(rbind,
            parallel::mclapply(
            X=z_range, 
            FUN=fisher_wrapper, 
            mc.cores = 6
            )
          )
    
  # Count the number of CpGs in the CTCF domains of interest
  tmp <- data.frame(
    cpgs.ix = pvals.ix,
    ctcf.ix = ctcf.ix
  ) %>% 
    group_by(ctcf.ix) %>% 
    summarize(N.CpG = n())
    
  # Pack the Fisher's p-value, CTCF start/stop into one df
  ctcfs.df <- cbind(
    ctcfs[z_range],
    data.frame(p.fisher = ctcf.p),
    N.CpG = tmp$N.CpG
    )
  
  # Write out data for use next time 
  save(list = c("ctcfs.df"), file = ofile)
  
} 
```


```{r}

ctcfs.df %>% 
  ggplot(aes(x = p.fisher)) +
  geom_histogram() +
  theme_minimal()

```


```{r}
ctcfs.df %>%
  dplyr::mutate(bin = cut(N.CpG,
                          breaks = c(0, 5, 10, 25, 75), 
                          right = T)) %>% 
  ggplot(aes(x = p.fisher, fill = bin)) +
  geom_histogram(bins = 50, aes(y = after_stat(density))) +
  facet_wrap(.~bin) +
  theme_minimal()
```
The significant CTCF binding-domains seem to be spread (roughly) by CTCF-binding domain size.

```{r}
sum(ctcfs.df$p.fisher < 0.01 / nrow(ctcfs.df))
```


# CTCF

> CTCF is known to bind a CpGrich consensus sequence that is usually unmethylated, as CTCF preferentially binds to unmethylated elements

-Kim et. al. 
