---
title: "CTCF Analysis"
author: "Coleman Breen"
date: "2022-11-28"
output: pdf_document
---

```{r}
library(tidyverse)
library(data.table)
library(poolr)
library(GenomicRanges)
library(ensembldb)
library(clusterProfiler)
library(EnsDb.Hsapiens.v86)

endb <- EnsDb.Hsapiens.v86
```

```{r}
pvals <- fread("../DataRaw/pvals.bed")
enhancers <- fread("../DataReference/enhancer_promoter_map.hg38.BED11") %>% select(-strand)
```


```{r}
pvals.gr <- makeGRangesFromDataFrame(pvals, 
                                     keep.extra.columns = T, 
                                     ignore.strand = T, 
                                     starts.in.df.are.0based = T
                                     )


enhancer.gr <- makeGRangesFromDataFrame(enhancers, 
                                        seqnames.field = "e.chr",
                                        start.field = "e.start",
                                        end.field = "e.end",
                                        keep.extra.columns = T, 
                                        ignore.strand = T, 
                                        starts.in.df.are.0based = T
                                        )

# Overlap
overlaps <- findOverlaps(pvals.gr, enhancer.gr)

```

# Do the p-values 

```{r}
ix <- queryHits(overlaps)

pvals[ix, ] %>% 
  ggplot(aes(x = p.corrected)) +
  geom_histogram() +
  theme_minimal()
```

Low p-value sites are somewhat absent from CTCF-binding domains. However, if there are, say, 10 out of 10 CpGs which are DMPs in one CTCF, that might mean there is something interesting nearby.

Show the p-values associated with one CTCF region (arbitrarily picked 71).

```{r}

pvals.ix <- queryHits(overlaps)
ref.ix <- subjectHits(overlaps)

fisher_wrapper <- function(z){
   # Which CpGs are in the zth ctcf
  keepix <- (ref.ix == z)
  pp <- pvals.gr[pvals.ix[keepix]]$p.corrected
  
  # Run Fisher's method to combine p-values
  fisher.out <- fisher(pp)
  fisher.out$p
}

z_range <- unique(ref.ix)

```

Now do the same thing for all 400k regions.

```{r}
# Run the Fisher routine on all the CTCF binding domains
enhancer.p <- 
  do.call(rbind,
          parallel::mclapply(
          X=z_range, 
          FUN=fisher_wrapper, 
          mc.cores = 6
          )
        )
```


```{r}
# Count the number of CpGs in the CTCF domains of interest
tmp <- data.frame(
  cpgs.ix = pvals.ix,
  ref.ix = ref.ix
) %>% 
  group_by(ref.ix) %>% 
  summarize(N.CpG = n())
    
# Pack the Fisher's p-value, CTCF start/stop into one df
out.df <- cbind(
  enhancers[z_range],
  data.frame(p.fisher = enhancer.p),
  N.CpG = tmp$N.CpG
  ) %>% 
  unite("e.ID", c(e.chr, e.start), sep = ":") %>%
  distinct(e.ID, .keep_all = T)
  
# Write out data for use next time 
save(list = c("enhancer_promoter.df"), file = ofile)

```


```{r}

p <- out.df  %>%
  ggplot(aes(x = p.fisher)) +
  geom_histogram(bins = 50) +
  theme_minimal() +
  ggtitle("Fisher's p-values for ~22k blood-specific enhancers")

cowplot::save_plot(filename = "../Figs/Enhancers/fisherCorrectedP.hist.png", p)

```

We see something similar with CTCF sites--bimodal. Correction needed? 

```{r}
sum(out.df$p.fisher < 0.01 / nrow(out.df))
```

606 significant enhancers at Bonferroni-adjusted level of 0.01.

# Annotate to genes

```{r}

# Map back to promoters
sig.IDs <- out.df %>% 
  dplyr::filter(p.fisher < 0.01 / nrow(.)) %>%
  dplyr::pull(ID)

sum(enhancers$ID %in% sig.IDs)

sig.promoters <- enhancers %>%
  dplyr::filter(ID %in% sig.IDs) %>%
  makeGRangesFromDataFrame(
    seqnames.field = "p.chr",
    start.field = "p.start",
    end.field = "p.end",
    strand.field = "p.strand"
  )


```

606 significant enhancers map to 609 significant promoters.

```{r}
genes <- genes(endb)
seqlevelsStyle(genes) <- "UCSC"

keepix <- nearest(sig.promoters, genes)

genes_list <- genes$gene_id[keepix]


go.out <- clusterProfiler::enrichGO(genes_list, 
                                    keyType = "ENSEMBL", 
                                    ont="ALL",
                                    OrgDb = "org.Hs.eg.db")
```

```{r}
data.frame(go.out) %>%
  arrange(p.adjust) %>%
  dplyr::select(c(Description, GeneRatio, p.adjust, ONTOLOGY)) %>%
  head(25) %>%
  ggplot(aes(x = Description, y = p.adjust, fill = ONTOLOGY)) +
  geom_col() +
  coord_flip() +
  theme_minimal() -> p

```

